### 2020-03-09 manuscript/manuscript.rmd
update manuscript and add senstive analyses

Phylogenomic studies based on hundreds to thousands of protein coding genes have greatly improved our understanding of evolutionary relationships in the animal tree of life. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions. We next quantify the  support of phylogenetic signal and conduct sensitive analyses in several representative datasets. We found that the Porifera-sister can only be recovered with a combined factors in certain data matrices, with inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories and more closely related outgroups, whereas Ctenophora-sister is strongly supported in most other cases. We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic. Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. IN this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT). If there are no modifiers, then it is implied that site homogeneous models are used. Moreover, the Gamma-rate heterogeinity is used in almost every analysis.

@@

High-throughput sequencing allows investigators interested in the root of animal phylogeny to assemble matrices with hundreds or thousands of genes (Table 1). Surprisingly, gene overlap between studies has rarely been assessed.  Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), animal root studies relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered. As a result, the genes selected for analysis can vary widely between studies.

@@

Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig. 1C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated-- Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.

@@

## Overview of published analyses
Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, including 1 data matrices from 12 studies (Table 1). These matrices include all relevant studies published before 2018 for which we could obtain data matrices with martition files that described boundaries between genes.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig. 1). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies only include less than five ctenophore species (Extended Data Fig. 1). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal taxa) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig. 2B). Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C). We found Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in all other cases.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 4; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive. The GTR+C60 models requires to estimate up to over 10,000 (e.g. GTR+C60 contains 234*60 parameters) free parameters inferred from matrices, whereas only several hundred parameters needs to estimate in site-homogeneous+C60 models. We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C models with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices, indicating that the support of Porifera-sister is increasing with inclusive of the more closely related outgroups and CAT models. These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups. More importantly, our results indicated that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several key data matrices.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C60 models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Whelan2017_full matrices.

@@

### Sensitive test of CAT categories {place holder for sensitive results}
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix. Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx). Although the true number of substitutional categories in empirical data matrices is impossible with current bioinformatic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results,  potentially indicating that the C60 models might be already overfitting to most matrices. These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses.

We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices. Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices.Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories. The major transition occurred at 90 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four choanimalia matrices. Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than site-homogeneous models.

The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model. Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups. Interestingly, we found both Whelan2017_full and Whelan2017_strict matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models.  
Thus, our results also strongly suggested that the exchangeability matrix of evolutionary model (a more simplistic Poisson model and more realistic GTR model) also has strong impact to the rooting position of animal phylogeny in PhyloBayes.

@@

## Future direction

Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

#### Sensitive analyses with different number of substitutional categories

>>>>

Phylogenomic studies based on hundreds to thousands of genes have greatly improved our understanding of evolutionary relationships in the animal tree of life. However, considerable debate still remains regarding the deepest split in the animal tree of life, with Ctenophora-sister and Porifera-sister emerging as the two primary alternative hypotheses. This has been one of the most difficult to resolve nodes in animal tree of life, and has major implications for our understanding of the earliest events in animal evolution. Here we systematically explore this question by synthesizing data and results from all previous phylogenomic analyses and perform new analyses with these data to characterize the differences between matrices under consistent conditions. Our reanalyses are generally consistent with previous results. We next quantify the support of phylogenetic signal and conduct sensitive analyses in several representative data matrices. We found that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions. We also show that recoding amino acid data into reduced state space, another approach that has been suggested to address variation in amino acid frequencies, is problematic. Thus, our results provide a comprehensive overview of the current understanding of the first diverging animal branch and why different studies yield different answers. This synthesis provides an integrative overview of the challenge and provides direction for future studies to address difficult phylogenetic problems.

@@

Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

@@

High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1). Surprisingly, gene overlap between studies has rarely been assessed. Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details). As a result, the genes selected for analysis can vary widely between studies.

@@

Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig. 1C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.

@@

## Overview of published analyses
Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 12 studies (Table 1, Table ./data_processed/tables/previously_published_analyses.tsv). These matrices include all relevant studies published before 2018 for which we could obtain data matrices with gene partitions that described boundaries between genes.

@@

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig. 1). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.

@@

We first used the Metazoa BUSCO genes [@waterhouse2017busco0] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of genome or transcriptomic assemblies [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig. 2B). Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also annotated each gene partition against SwissProt [@boeckmann2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.

@@

Here we summarized all the `r nrow(analyses_published)` previous phylogenetic analyses from `r length(unique(analyses_published$manuscript))` studies (Fig. 3 and Supplementary Table 2). Among these studies, the main conclusions of five studies strongly in favor of Porifera-sister and 10 in support of Ctenophora-sister (Table 1). Moreover, three studies are secondary studies without any new data matrices. Interestingly, we found that all analyses strongly support of Ctenophora-sister once site-homogeneous models were used, no matter what outgroup choice, phylogenetic inference and whether data-recoding are used. The support for Porifera-sister increases with the exclusion of more distantly related outgroups and with the use of site-heterogeneious CAT (Poisson+CAT or GTR+CAT) models in several matrices, including Philippe2009, Ryan2013_est and Whelan2015. It should be noted that many BI analyses with CAT model are not well resolved (either not reach convergence or do not provide significant support for the key node of animal evolution) mainly due to computational limits of CAT models employed in PhyloBayes. The strong support of Porifera-sister can only be recovered from analyses using Poisson+CAT and recoding methods with GTR+CAT models in several matrices. Although GTR+CAT model generally has better model adequacy than Poisson+CAT or site-homogeneous models from previous studies[@pisani2015genomic; @feuda2017improved], Porifera-sister is not always supported since several analyses strongly in favor Ctenophora-sister even with only choanofalgellates are used as outgroups (Fig. 3). Thus, our summary of previous phylogenetic analyses strongly indicated that the trend of the support of Porifera-sister is generally increasing with the use of more complex CAT models and more closely related outgroups in several representative matrices, whereas Ctenophora-sister is supported in most other conditions.

@@

One of the challenges of interpreting support for the placement of the animal root is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Thus, it easily causes confusions with the increase number of studies.  Here we first reanalyze the primary matrices from each study under consistent conditions with IQtree [@nguyen2014iq] under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood (ML) framework [@zhou2017evaluating] (Fig. 4; Supplementary Table 3). Importantly, we also include a CAT-like, site-heterogeneous C10 -C60 (C60) models [@si2008empirical] that are implemented in both ML framework in our analyses and it is interesting to compare results between C60 model to CAT models since it has never been used in any matrices analyzed here.

Here we reanalyzed 17 matrices from `r length(unique(analyses_published$manuscript))` studies that are used as main conclusions from their original paper (Table 1). We also progressively trimmed each data matrix with three level of outgroup choice if applicable (Fig 1C). We first tested a variety of models for each matrix and compared the relative fit of site-homogeneous and site-heterogeneous C60 models using ModelFinder [@kalyaanamoorthy2017modelfinder] in IQtree (except only LG+C60 models are used in 3 large matrices). We found in all cases, WAG/LG+C60 models fit these matrices better than the site-homogeneous models under BIC criteria and we then inferred support under the best-fit model from IQtree (Extended Data Table 1). It should be noted that we didn't include GTR+C60 models in model testing since it is computational too expensive with large number of parameters that need to be estimated.  We then analyzed every matrix under a panel of standard site-heterogeneous and site-homogeneous models, including WAG, GTR and Poisson+C60. Moreover, we also used the best-fit model identified above with the removal of all distantly related outgroups and only include Choanoflagellata as outgroups. Interestingly, with the exception of Moroz2014_3d and all Nosenko2013 matrices that supports neither Porifera-sister or Ctenophora-sister, all analysis conducted herein in IQtree strongly supported the ctenophore-sister hypothesis, even with CAT-like, site-heterogeneous C60 models (Poisson+C60 and WAG/LG+C60) with the most closely related outgroups (Fig. 4; Supplementary Table 3).

@@

Interestingly, we found strong supports for both hypotheses when using Poisson-CAT model in analyzed Choanimalia matrices. Similar to ML analyses, we found no strong support for either hypothesis in Moroz2013_3d and Nosenko2013 matrices. Consistent with previous study [@Philippe:2009hh; @pisani2015genomic], we recovered strong support of Porifera-sister in matrices of Philippe2009, Ryan2013_est and Whelan2015. Interestingly, we also found a strong support of Porifera-sister in two Whelan2017 matrices[@whelan2017ctenophore] and a weaker support of subsampled matrix of Simion2017 [@simion2017large]. The Ctenophora-sister were strongly supported in all other matrices. For the matrices have conflicted results between Poisson-CAT and models employed in IQtree, we further ran Poisson+CAT models with inclusion of more distantly related outgroups to evaluate how outgroup choice may affect phylogenetic inference under CAT model. We found a weak support of Porifera-sister in all Holozoa matrices and strong support of Ctenophora-sister in all analyzed Opisthokonta matrices. These results suggest that Porifera-sister can only be recovered in certain gene sampling in several representative matrices with the exclusive of more distantly related outgroups when analyzed using Poisson-CAT. Thus, our results from reanalyses are largely similar to previous studies and indicate that both model choice and outgroup sampling play a major role in incongruent results of animal origin in several representative data matrices.

@@

In contrast, the phylogenetic signal of site-heterogenous models changed dramatically compared to the site-homogenous models. Although Ctenophora-sister also had the higher proportions of supporting genes with C60 models, we found that the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models, especially in matrices from Ryan2013_est that nearly 30% of genes changed from strong to week signal (\Delta$lnL<2) (Fig. 5B; Supplementary Table 4). It is not clear why the phylogenetic signal decreases in many genes with a better fitting model as suggested by model testing in IQtree, but our results indicated the C60 models lack of phylogenetic signal compared to site-homogeneous models in the node of the root position of animal phylogeny. In contrast to the C60 models, the phylogenetic signal largely increased in Poisson-CAT models and more importantly, the outgroup choice has a major effect of the distribution of phylogenetic signal (Figs. 5A, C). We found that support for Ctenophora-sister largely decreases and Porifera-sister increases (~ 20% of sampled gene partitions) when exclusive of more distantly related outgroups with the Poisson-CAT models (Supplementary Table 4). The results here are further corroborated by our phylogenomic analyses in BI, which the support of Porifera-sister generally increases in Holozoa and Choanimalia matrices.

We also examined the level of congruence of gene partitions that are supported in each hypothesis across different analyses in Whelan2017_full matrices (Fig. Extended Data Fig. 3). Consistent with the summary of \Delta$GLS\ analysis, We found gene population that supports in either hypothesis are largely congruent between site-homogenous and C models, whereas little overlap was observed in Poisson-CAT models (especially with different outgroup choice). In other words. our results strongly indicated that the distribution of phylogenetic signal of animal origin is largely sensitive to model selection (site-homogeneous vs. site-heterogeneous models). More importantly, the site-heterogeneous CAT model is also more sensitive to the outgroup choice than site-homogeneous and C60 models in Phillippe2009 and Whelan2017_full matrices.

@@

### Sensitive test of CAT categories {place holder for sensitive results}
The major difference between site-heterogeneous CAT and C60 models are the number of substitutional categories that are used in different data matrices. CAT models employ a Dirichlet process prior to infer number of substitutional categories or profiles with different nucleotide or amino acid frequencies to account for substitutional heterogeneity across a phylogenetic data matrix [@Lartillot:2004dq] . Here, we reported the number of categories inferred from CAT model for each PhyloBayes analysis (Table XX). The number of categories is considerable larger than 60 categories used in C60 models, and in most cases even larger than the gene partitions within each data matrix (Table xx). Although the true number of substitutional categories in empirical data matrices is impossible to determine with current phylogenetic tools, we found that the amino frequencies of several categories inferred by C60 model are close to zero from IQtree results, potentially indicating that the C60 models might be already overfitting to most matrices. These results are consistent with the findings that CAT model often overestimate the true number of substitutional categories, especially for large data matrix from simulation analyses [@whelan2016let].

We next explore how the increase of number of categories may affect the phylogenetic inference related to the animal phylogeny with a panel of different substitutional categories (nCAT = 60, 90, 120, 150, 180) in the representative data matrices. Surprisingly, we found Poisson+nCAT60 showed exact same topology as Poisson+C60 or WAG+C60 with strong support of Ctenophora-sister, suggesting that the different phylogenetic analysis software (IQtree and PhyloBayes) and phylogenetic inference (ML and BI) has limited impact on phylogenetic inference to these matrices with similar models. Importantly, our results gave decreased support for Ctenophora-sister and increased support of Porifera-sister with the increase of number of substitutional categories. The major transition occurred at 60 – 120 categories, and the support for Ctenophora-sister disappears after 120 categories in the four representative Choanimalia matrices. Interestingly, the results of our sensitive analyses further corroborated by the recent simulation study that the tendency of overestimate the true number of categories of Poisson-CAT or GTR-CAT model can potentially lead to a less accurate tree than the site-homogeneous models [@whelan2016let].

The GTR-CAT model often has a better model-fitting than Poisson-CAT model in previous analyses, and it is also suggested a better performance than Poisson-CAT model. Due to computational efficiency, we only ran GTR-CAT and GTR-nCAT60 models in the representative data matrices with the most closely related choanofalgellates as outgroups. Interestingly, we found both Whelan2017_full and Whelan2017_strict matrices shifted from Porifera-sister with Poisson-CAT model to Ctenophora-sister using GTR-CAT model. Moreover, we also found all results strongly supported the Ctenophora-sister with GTR-nCAT60 models. Thus, these results suggested that the Porifera-sister can only be recovered with several combined factors, including inclusive of a more simplistic evolutionary model, an unrealistic large number of substitution categories, more closely related outgroups and certain gene samplings, whereas Ctenophora-sister is supported in most other conditions.

@@

## Future direction
Many key positions of animal phylogeny can be reconstructed by increasing genome-scale data within a phylogenomic framework, although new challenges have also arised. Overcoming the challenges to inferring the earliest diverging animal lineage is particular difficult and here we discuss the several aspects that will help to improve our understanding of animal evolution. First, a broader sampling of high-quality sponge and ctenophore genomes are needed to shorten the long branch since they are among the longest branches in the animal phylogeny (currently only 2 ctenophore and 3 sponge genomes are available). Second, accurate reconstruction of orthologs and a consensus analytical strategy should be formalized with investigators to improve communications and reproducibility of phylogenetic workflow [@guang2016integrated]. Last, we also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop and evaluate methods or evolutionary models to address this difficult phylogenetic problem. More importantly, we also hope all investigators critically evaluate the competing hypotheses without bias instead of showing preferences to one hypothesis over as a priori due to historical reasons or previous work. Finally, other sources of evidence may also provide important insight to animal origin, including morphological characters [@mah2014choanoflagellate; @nielsen2019early] and fossil records [@zhao2019cambrian]. We hope the future breakthroughs that will allow us to move the field forward and a consensus viewpoint of early animal evolution will be reached in the coming years.

@@

#### Sensitive analyses with different number of substitutional categories
To explore how the number of categories used in CAT model
